The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex
Small changes throughout

\subsection{Effect of Midplane Offset} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate or erroneous determination of the midplane leads to a phase-dependence of the actions
calculated from an orbit specified by a phase-space point and an assumed
potential model. The $y$-axis corresponds to the vertical height of the orbit
and the $x$-axis shows orbital phase. The solid gray curve indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is the midplane location in the potential
model used by an observer to integrate the orbit of the star and hence
estimate its actions, which is offset from the true midplane of the star's
orbit. The model potential is otherwise identical to the one in which the star
is actually moving.

@@

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g. the midplane or
Galactic center distance) or velocity (e.g. the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

@@

\subsection{Numerical Methods} \label{ssec:action_comp}
We now quantify the argument made in Section~\ref{ssec:cartoon} using
numerical computations of the actions for a range of orbits in a model
Galactic potential. We compute actions as in \citet{2018ApJ...867...31B},
using the code \texttt{gala} v0.3 to perform orbit integrations and conversion
to action space \citep{2017JOSS....2..388P,Price-Whelan:2018}. To compute
actions we use the torus-mapping technique first presented by
\citet{1990MNRAS.244..634M} and adapted by \citet{2014MNRAS.441.3284S} to
calculate actions for an orbital time-series starting from a phase-space
position $(x, v)$ and integrated in a potential $\Phi$. We use the default
\texttt{MWPotential} as our potential, which is based on the Milky Way
potential available in \texttt{galpy} \citep{2015ApJS..216...29B}. This
potential includes a Hernquist bulge and nucleus \citep{1990ApJ...356..359H},
a Miyamoto--Nagai disk \citep{1975PASJ...27..533M}, and an NFW halo
\citep{1997ApJ...490..493N}, and is fit to empirically match some
observations. We use the Dormand-Prince 8(5,3) integration scheme
\citep{Dormand80:integrator} with a timestep of $1\,\Myr$ and integrate for
$5\,\Gyr$, corresponding to $\sim 20$ orbits for a Sun-like star.

@@

We consider that the observer makes a measurement every $\Myr$. At
each measurement, we compute the actions as outlined above using the true
potential but by specifiying the star's starting position using the
systematically offset coordinate system. Essentially we are shifting the
integrated orbit and then reintegrating at each point along the shifted orbit. The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first $\Gyr$ of the
orbit in the upper panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

We also perform the same procedure but assume a $100\,\pc$ offset in the $x$
component (i.e. subtracting the vector $(100, 0, 0)\,\pc$) in the lower
panels. This is equivalent to a systematic offset in the determination of the
distance from the Sun to the Galactic center.

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
appear as time-dependence in the actions, oscillating around
their value using the correct coordinate system. This time dependence comes even though
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality, we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed at different orbital phases for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

@@

The upper row of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread
induced in each action for an offset in the $z$-component (i.e. the Galactic
midplane). In the lower row we consider offsets in the $x$ component (i.e. the
Solar radius). The left, center, and right columns show the fractional spread
in the values computed for $J_z$, $J_{\phi}$, and $J_R$, respectively. We
compute the fractional spread $\Delta J_i/J_i$, where $\Delta J_i$ is defined
in the previous paragraph and $J_i$ is the action value computed in the true
coordinate system, over the course of the first $\Gyr$.

In the upper middle panel of Figure~\ref{fig:many_orbit_wrong_ref}, there is
essentially no spread in the determination of $J_{\phi}$. This is expected
since $J_{\phi}$ is independent of $z$ and is thus unaffected by offsets in
$z$, as discussed earlier. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for the thick-disk orbit holds for all
orbit types. This is also a demonstration of the robustness of the orbit
integration and action calculation methods we use.

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For the thin-disk orbit, a systematic offset of $\sim15\,\pc$ in the
$z$ coordinate already results in $25\%$ deviations in the actions. The
required offset for $25\%$ deviation is $\sim120\,\pc$ for the thick-disk
orbit. The halo orbit is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to a midplane offset of $500\,\pc$.

@@

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not necessarily due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar in mass to
Sagittarius, which has punched through the galactic disk outside the solar
circle, leaving behind some of its stars and inducing warping in the disk.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider
results from a live N-body simulation of a Sagittarius-like encounter. This
simulation offers us the ability to see how the midplane varies in a more
controlled environment. The details of the simulation are given in
\citet{2018MNRAS.481..286L}, but we briefly summarize the most relevant
details here. For the Milky Way, the dark halo is modeled as a Hernquist
sphere of mass $10^{12}\,\Msun$ and scale length of $52\,\kpc$, the disk
modeled as an exponential disk with a scale radius of $3.5\,\kpc$, scale
height $0.53\,\kpc$, and mass $6\times10^{10}\,\Msun$, and a Hernquist bulge
of mass $10^{10}\,\Msun$ and scale length $0.7\,\kpc$
\citep{1990ApJ...356..359H}. We use the L2 model, for which Sagittarius is
modeled as a dark matter Hernquist sphere of mass $8\times10^{10}\,\Msun$ and
scale length $8\,\kpc$, along with a stellar component modeled as a Hernquist
sphere of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All
components are realized with live distributions of N-body particles; the Milky
Way and Sagittarius are each initialized to be in equilibrium in isolation.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} (left,
center, and right panels) as a function of azimuthal angle. All calculations
are done at cosmological redshift $\z =0$. The local midplane is determined at
a position $\phi$ by taking the median height of all stars within
$R=0.5\,\kpc$ and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is
performed again using the new height $10$ times to converge on the local
midplane height. In order to allow for the possibility that the fiducial
galactocentric coordinate system is incorrect, we subtract the best fit
$A\sin{(\phi+B)}+C$ curve from each panel --- this figure is reproduced with
the original midplane determination (i.e. before subtracting the best fit sine
curve) in Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on
all stars within a $2\,\kpc$ height of the fiducial midplane to determine
$1\,\sigma $ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur
which bring the Sagittarius-like object closer to the Solar circle, we can see
that stronger midplane variations result. In fact, one can see that an $m=2$
mode is excited in the lower left panel at $t=6.0\,\Gyr$ --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited ($m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The fact that the $t=6.9\,\Gyr$
panel looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) suggests that midplane variations are
produced at least in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

To account for the effect of particle noise, we bootstrap resample stars
within a cylinder of stars with height $2\,\kpc$ and the same radius $1000$
times and determine the $1\,\sigma$ error bars by repeating the midplane
determination with those reselection of stars.

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane our imaginary observer would determine as a function of azimuth for
each galaxy, using their local ``solar neighborhood'' (the cylinder defined
above). We consider $50$ equally spaced bins in azimuth, a sufficient number
such that each cylinder shares no stars with its neighboring cylinder. The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the Solar circle spans
$185$, $162$, $84\,\pc$ for \mi{}, \mf{}, and \mm{}. In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the Solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to $\pm 50\,\pc$.

@@

In the upper panels, we see that the midplane is relatively flat in the inner
galaxy, but additional encounters drive strong midplane variation. In the
lower left panel, we see a strong $m=2$ mode develop, consistent with the
$R=8\,\kpc$ panel of Figure~17 in \citet{2018MNRAS.481..286L}. The lower right
panel, which shows the galaxy after some relaxation has occured is
qualitatively similar to the midplane variations we saw in the FIRE
simulations (Figure~\ref{fig:midplane}), evidence that they are at least
partially driven by mergers.

@@

Systematic variations in $v_z$ were first noted as asymmetries in the local
velocity distribution towards the North and South Galactic Caps from the
radial velocity surveys of the Sloan Digital Sky Surveys
\citep{2012ApJ...750L..41W} and RAdial Velocity Experiment
\citep{2013MNRAS.436..101W}. Subsequently, \citet{2013ApJ...777L...5C} pointed
out suggestions of an oscillation in average vertical velocities of order
$5\,\kms$  on $\sim\,\kpc$ scales looking toward the Galactic anticenter. Work
by the \textit{Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of these oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

>>>>

\subsection{Effect of Midplane Offset} \label{ssec:cartoon}
We present a cartoon approximation of an orbit in Figure~\ref{fig:cartoon} to show how an
inaccurate or erroneous determination of the midplane leads to a phase-dependence of the actions
calculated from a phase-space point and an assumed
potential model. The $x$-axis corresponds to orbital phase
and the $y$-axis to vertical height. The solid gray curve indicates the
``true'' orbit of the star; that is, as it oscillates around the
true midplane.

The dashed gray line, offset from the true midplane, is the midplane location
used by an observer to integrate the orbit of the star and estimate its
actions. The model potential is otherwise identical to the one in which the
star is actually moving.

@@

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g. the midplane or
Galactic center distance) or velocity (e.g. the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of the thin disk and a good description of the thick disk in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

@@

\subsection{Numerical Methods} \label{ssec:action_comp}
We now quantify the argument made in Section~\ref{ssec:cartoon} using
numerical computations of the actions for a range of orbits in a model
Galactic potential. We compute actions as in \citet{2018ApJ...867...31B},
using the code \texttt{gala} v0.3 to perform orbit integrations and conversion
to action space \citep{2017JOSS....2..388P,Price-Whelan:2018}. To compute
actions we use the torus-mapping technique first presented by
\citet{1990MNRAS.244..634M} and adapted by \citet{2014MNRAS.441.3284S} to
calculate actions for an orbital time-series starting from a phase-space
position $(x, v)$ and integrated in a potential $\Phi$. For our Galactic potential we use \texttt{MWPotential}, based on the Milky Way
potential available in \texttt{galpy} \citep{2015ApJS..216...29B}, which includes a Hernquist bulge and nucleus \citep{1990ApJ...356..359H},
a Miyamoto--Nagai disk \citep{1975PASJ...27..533M}, and an NFW halo
\citep{1997ApJ...490..493N}, and is fit to empirically match some
observations.
We use the Dormand-Prince 8(5,3) integration scheme
\citep{Dormand80:integrator} with a timestep of $1\,\Myr$ and integrate for
$5\,\Gyr$, corresponding to $\sim 20$ orbits for a Sun-like star.

@@

We consider that the observer makes a measurement every $\Myr$. At
each measurement, we compute the actions as outlined above using the true
potential but by specifiying the star's starting position using the
systematically offset coordinate system. Essentially we are shifting and then reintegrating at each point along the original orbit.
The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first $\Gyr$ of the
orbit in the upper panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

We also perform the same procedure in the lower panels but assume an $x$
component offset of $100\,\pc$, i.e., subtracting the vector $(100, 0,
0)\,\pc$. This is equivalent to a measurement error in the distance from the
Sun to the Galactic center.

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time/orbital phase at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
appear as time-dependence in the actions, oscillating around
their value using the correct coordinate system. This time dependence comes even though
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality, we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed at different orbital phases for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the true value (computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

@@

The upper row of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread
induced in each action for an offset in the $z$-component. In the lower row we consider offsets in the $x$ component (i.e. the
Solar radius). The left, center, and right columns show the fractional spread
in the values computed for $J_z$, $J_{\phi}$, and $J_R$, respectively. We
compute the fractional spread $\Delta J_i/J_i$, where $\Delta J_i$ is defined
in the previous paragraph and $J_i$ is the action value computed in the true
coordinate system, over the course of the first $\Gyr$.

In the upper middle panel of Figure~\ref{fig:many_orbit_wrong_ref}, there is
essentially no spread in the determination of $J_{\phi}$. This is expected
since $J_{\phi}$ is independent of $z$ and is thus unaffected by offsets in
$z$, as discussed earlier. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for the thick-disk orbit holds for all
orbit types. This is also a demonstration of the robustness of the
integration and action calculation methods we use.

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For the thin-disk (thick-disk) orbit, a systematic offset of $15\,\pc$ ($120\,\pc$) in the $z$-coordinate results in a $25\%$ deviation in $J_z$.
The halo orbit is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to an offset of $500\,\pc$.

@@

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not necessarily due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar in mass to
Sagittarius, which has punched through the Galactic disk outside the solar
circle, leaving behind some of its stars and inducing warping in the disk.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider
results from a live N-body simulation of a Sagittarius-like encounter. This
simulation offers us the ability to see how the midplane varies in a more
controlled environment. The details of the simulation are given in
\citet{2018MNRAS.481..286L}, but we briefly summarize the most relevant
details here.

For the Milky Way, the dark halo is modeled as a Hernquist
sphere of mass $10^{12}\,\Msun$ and scale length of $52\,\kpc$, the disk
modeled as an exponential disk with a scale radius of $3.5\,\kpc$, scale
height $0.53\,\kpc$, and mass $6\times10^{10}\,\Msun$, and a Hernquist bulge
of mass $10^{10}\,\Msun$ and scale length $0.7\,\kpc$
\citep{1990ApJ...356..359H}. We use the L2 model, for which Sagittarius is
modeled as a dark matter Hernquist sphere of mass $8\times10^{10}\,\Msun$ and
scale length $8\,\kpc$, along with a stellar component modeled as a Hernquist
sphere of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All
components are realized with live distributions of N-body particles; the Milky
Way and Sagittarius are each initialized to be in equilibrium in isolation.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} (left,
center, and right panels) as a function of azimuthal angle, at cosmological redshift $\z =0$. The local midplane is determined at
a position $\phi$ by taking the median height of all stars within
$R=0.5\,\kpc$ and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is
performed again using the new height $10$ times to converge on the local
midplane height. In order to allow for the possibility that the fiducial
galactocentric coordinate system is incorrect, we subtract the best fit
$A\sin{(\phi+B)}+C$ curve from each panel. We then bootstrap resample $1000$ times on
all stars within a $2\,\kpc$ height of the fiducial midplane to determine
$1\,\sigma $ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$, with an encounter happening close to the Solar circle near $t=6.0\,\Gyr$. The fact that the $t=6.9\,\Gyr$ panel, which shows the midplane variation after some relaxation,
looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that midplane variations are
generated, in part, by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

To account for the effect of particle noise, we bootstrap resample stars
within a cylinder of height $2\,\kpc$ and the same radius $1000$
times and determine the $1\,\sigma$ error bars by repeating the midplane
determination with those reselection of stars.

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane an observer would determine as a function of azimuth for
each galaxy, using their local ``solar neighborhood'' (the cylinder defined
above). We consider $50$ equally spaced bins in azimuth, a sufficient number
such that each cylinder shares no stars with its neighboring cylinder. The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the Solar circle spans
$185$, $162$, $84\,\pc$ for \mi{}, \mf{}, and \mm{}. In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the Solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to $\pm 50\,\pc$.

@@

In the upper panels, we see that the midplane is relatively flat in the inner
galaxy, but additional encounters drive strong midplane variation. In the
lower left panel, we see a strong $m=2$ mode develop, consistent with the
$R=8\,\kpc$ panel of Figure~17 in \citet{2018MNRAS.481..286L} ($m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The lower right
panel, which shows the galaxy at $t=6.9\,\Gyr$ when some relaxation has occured, is
qualitatively similar to the midplane variations we saw in the FIRE
simulations (Figure~\ref{fig:midplane}), evidence that they are at least
partially driven by mergers.

@@

Systematic variations in $v_z$ and number were first noted as asymmetries in the local
velocity distribution towards the North and South Galactic Caps from the
radial velocity surveys of the Sloan Digital Sky Surveys
\citep{2012ApJ...750L..41W} and RAdial Velocity Experiment
\citep{2013MNRAS.436..101W}. Subsequently, \citet{2013ApJ...777L...5C} pointed
out suggestions of an oscillation in average vertical velocities of order
$5\,\kms$  on $\sim\,\kpc$ scales looking toward the Galactic anticenter.

Work
by the \textit{Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of these oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}, which can be explained with models of Sagittarius-like encounters \citep{2013MNRAS.429..159G,2018MNRAS.481..286L,2018arXiv180800451L}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.
